---
title: "Mini Project 02"
author: "Lasantha Kulasooriya(S/17/404)"
date: "2023-07-28"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Canonical Correlation Analysis

### 1. Data Preprocessing

```{r}
#import data set
data <- read.csv('../data/bodyPerformance.csv',skip=1, col.names = c("Age", "Gender","Height", "Weight", "Bodyfat", "Dialostic","Systolic", "Gripforce", "SitandBend", "Situps", "Broadjumps", "Class"))
head(data)
```

```{r}
#dimensions
dim(data)
```

-   The data set contains 13393 rows in 12 columns.

```{r}
#structure of the dataset
str(data)
```

-   Data set contains two character variables and ten numerical variables.

```{r}
#missing values
library(mice)
```

```{r}
miss.count <- function(x){
  sort(colSums(is.na(x)), decreasing = TRUE )
}
miss.count(data)
```

```{r}
md.pattern(data, rotate.names = TRUE, plot = TRUE)
```

-   Data set contains no any missing values.

```{r}
#summary of the data set
summary(data)
```

```{r}
library(dplyr)
data<- data %>% 
    select( -Gender, -Dialostic,-Systolic, -Class)
```

```{r}
head(data)
```

-   Four physiological and Four exercise variables are measured by Korean sports promotion foundation.\
    Physiological variables (Y) : Age , Height , Weight , Body fat percentage\
    Exercise variables (X) : Grip Force, Sit and bend forward , Sit-ups , Broad jump

```{r}
#divide the data set in to two sets as physiology and exercise by combining the relevant variables.

physiology.set <- select(data, Age, Height, Weight, Bodyfat)
exercise.set <- select(data, Gripforce, SitandBend, Situps, Broadjumps)
```

```{r}
#pshciology
head(physiology.set)
```

```{r}
#exercise
head(exercise.set)
```

```{r}
library(ggplot2)
library(tidyr)
library(gridExtra)

physiology_long <- pivot_longer(physiology.set, cols = everything(), names_to = "physiology", values_to = "value")
exercise_long <- pivot_longer(exercise.set, cols = everything(), names_to = "exercise", values_to = "value")

# boxplot for physiology set
physiology_plot <- ggplot(physiology_long, aes(x = physiology, y = value, fill = physiology)) +
  geom_boxplot() +
  labs(title = "Physiological variables", x = "Variables", y = "Value") +
  scale_fill_manual(values = c("red", "green", "royalblue2","yellow")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

# boxplot for exercise set
exercise_plot <- ggplot(exercise_long, aes(x = exercise, y = value, fill = exercise)) +
  geom_boxplot() +
  labs(title = "Exercise variables", x = "Variables", y = "Value") +
  scale_fill_manual(values = c("red", "sienna", "palevioletred1", "royalblue2")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")


grid.arrange(physiology_plot, exercise_plot, ncol = 2)

```

-   There are so many outliers in the dataset.

```{r}
library(CCA)
cormat=matcor(exercise.set,physiology.set)
cormat
```

```{r}
img.matcor(cormat, type=2)
```

-   Among the exersice set(X) highest positive correlation exists between the situps and broadjumps. And also between the gripforce and sitandbend exersices there is low negative correlation.

-   Among the physiology set(Y) highest negative correlation exists between Age and weight.

-   When considering the correlation between X set of variables with Y set of variables, among most of them have only moderate positive correlations.

Before applying the canonical correlation analysis, we need to check the multivariate normality of data.

### 2. Test the Assumption of Normality

```{r}
library(MVN)
# Madia test for multivariate normality
mvn(data = data,subset = NULL,mvnTest = "mardia")
```

```{r}
# Henze-Zirkler's test for multivariate normality
mvn(data = data,subset = NULL,mvnTest = "hz")
```

-   According Mardia's test and Henze-Zirkler's test, we can observe that the multivariate normality is not there. By Anderson-Darling test we can observe that all the variables are non-normal. The reason may be the outliers.

```{r}
result<-mvn(data = data,multivariateOutlierMethod = "adj",
showOutliers = TRUE, showNewData = TRUE)
```

-   There are 923 outliers and the observation 3355 and 2657 are extreme outliers.

```{r}
#remove outliers using standarization 
remove_outliers<- function(df, cols = names(df)) {
  for (col in cols) {
    col_mean <- mean(df[[col]])
    col_std <- sd(df[[col]])
    df[[col]] <- (df[[col]] - col_mean) / col_std
  
    threshold <- 2
    
    df <- df[abs(df[[col]]) <= threshold, ]
    df[[col]] <- df[[col]] * col_std + col_mean
  }
  
  return(df)
}
```

```{r}
data.new <- remove_outliers(data)
```

```{r}
dim(data.new)
```

```{r}
# mardia test for multivariate normality
mvn(data = data.new,subset = NULL,mvnTest = "mardia")
```

```{r}
# Henze-Zirkler's test for multivariate normality
mvn(data = data.new,subset = NULL,mvnTest = "hz")
```

Outliers removed data also violates the multivariate normality assumption.

```{r}
# Anderson-Darling test to check univariate normality
mvn_un <- mvn(data = data.new, univariateTest = "AD" , desc=T)
mvn_un$univariateNormality
```

According to the above tests we confirms that all the variables are non-normal. Transforming data to achieve univariate normality.

```{r}
data.new[data.new <= 0] <- 0.001   

```

```{r}
#Log transformation
data.log= log(data.new)
```

```{r}
# Anderson-Darling test to check univariate normality
mvn_un.log <- mvn(data = data.log, univariateTest = "AD" , desc=T)
mvn_un.log$univariateNormality
```

```{r}
# mardia test for multivariate normality
mvn(data = data.log,subset = NULL,mvnTest = "mardia")
```

```{r}
# Henze-Zirkler's test for multivariate normality
mvn(data = data.log,subset = NULL,mvnTest = "hz")
```

Both multivariate and univariate noramality assumptions are violated. Now by assuming multivariate normality.

### 3. Canonical Correlations

```{r}
str(data.new)
```

```{r}
library(dplyr)
physiology.set.new <- data.new[,1:4]
exercise.set.new <- data.new[,-(1:4)]
```

```{r}
# canonical correlations
library(CCP)
cca.raw <- cancor(physiology.set.new, exercise.set.new)
cca.raw
```

### 4. Normalized Canonical Correlations

```{r}
# normalized canonical correlations
cca.scaled<-cancor(scale(physiology.set.new, scale=T, center=T),scale(exercise.set.new, scale=T,center=T))
cca.scaled
```

-   The canonical correlations are the same, but the coefficients of canonical variates are normalized to have Var(U) = 1 and Var(V) = 1.

### 5. Test the significance of Canonical Correlations

```{r}
dim(data.new)
```

```{r}
rho <- cca.scaled$cor
# Calculate p-values using the F-approximations of different test statistics:
p.asym(rho,10543,4,4,tstat="Wilks")
```

```{r}
p.asym(rho,10543,4,4, tstat = "Hotelling")
```

```{r}
p.asym(rho,10543,4,4, tstat = "Pillai")
```

```{r}
p.asym(rho, 10543,4,4, tstat = "Roy")
```

```{r}
# Plot the F-approximation for Wilks' Lambda, considering 3, 2, or 1 canonical correlation(s):
res1 <- p.asym(rho,10543,4,4)
```

```{r}
plt.asym(res1,rhostart=1)
```

```{r}
plt.asym(res1,rhostart=2)
```

```{r}
plt.asym(res1,rhostart=3)
```

```{r}
plt.asym(res1,rhostart=4)
```

-   All the 4 canonical correlations are significant at 5% level of significance.

### 6. Canonical Correlation Plots

#### 6.1 First Canonical Correlation Plot

```{r}
xx<-scale(physiology.set.new,scale=T,center=T)
yy<-scale(exercise.set.new,scale=T,center=T)
scorex<-xx%*%cca.scaled$xcoef[,1]
scorey<-yy%*%cca.scaled$ycoef[,1]
plot(scorex,scorey,type="n")
text(scorex,scorey,row.names(data.new),cex=.6)
```

#### 6.2 Second Canonical Correlation Plot

```{r}
xx<-scale(physiology.set.new,scale=T,center=T)
yy<-scale(exercise.set.new,scale=T,center=T)
scorex<-xx%*%cca.scaled$xcoef[,2]
scorey<-yy%*%cca.scaled$ycoef[,2]
plot(scorex,scorey,type="n")
text(scorex,scorey,row.names(data.new),cex=.6)
```

#### 6.3 Third Canonical Correlation Plot

```{r}
xx<-scale(physiology.set.new,scale=T,center=T)
yy<-scale(exercise.set.new,scale=T,center=T)
scorex<-xx%*%cca.scaled$xcoef[,3]
scorey<-yy%*%cca.scaled$ycoef[,3]
plot(scorex,scorey,type="n")
text(scorex,scorey,row.names(data.new),cex=.6)
```

#### 6.4 Fourth Canonical Correlation Plot

```{r}
xx<-scale(physiology.set.new,scale=T,center=T)
yy<-scale(exercise.set.new,scale=T,center=T)
scorex<-xx%*%cca.scaled$xcoef[,4]
scorey<-yy%*%cca.scaled$ycoef[,4]
plot(scorex,scorey,type="n")
text(scorex,scorey,row.names(data.new),cex=.6)
```

### 7. Canonical Loadings

```{r}
cancor<-cc(physiology.set.new, exercise.set.new)
comput(physiology.set.new, exercise.set.new,cancor)[3:6]
```
